Haul-out behavior and aerial survey detectability of seals in the Bering and Chukchi seas

### 2021-12-23 analysis/paper/London_HauloutBeahvior.Rmd
ADDRESS PCONN INITAL COMMENTS to v1.5

For ice-associated seals, haul-out behavior is tightly linked with sea ice.
Ribbon seals (_Histriophoca fasciata_) haul out of the water almost exclusively
on sea ice and are mostly pelagic outside the spring pupping, breeding, and
molting season [@boveng2018a]. While spotted (_Phoca largha_) and bearded
(_Erignathus barbatus_) seals rest on coastal features, they strongly prefer sea
ice as a resting platform during the spring and early summer [@frost2018].
Ringed seals (_Phoca hispida_) --- not included in this study --- bask on
sea-ice but also within snow lairs built during the spring. The remoteness of
the Bering and Chukchi seas mean direct scientific observation of seal behavior
is impractical. Thus, bio-logging devices are, thus, especially useful tools for
collecting key information on movement and haul-out behavior for these species.

@@

While analyzing haul-out patterns clearly has utility for understanding natural
history and ecology, some studies have focused, as we do here, specifically on
quantifying the probability of seals being hauled out as a means of quantifying
seals in the water that are not counted during aerial surveys. These studies
typically use logistic regression-style analyses to estimate the time-specific
probability of being hauled out based on 'wet/dry' data relayed by bio-loggers.
In these models, haul-out probabilities were expressed as a function of
predictive covariates, such as time-of-day, day-of-year, sex, age class, and
environmental conditions (e.g., [@reder2003], [@bengtson2004], [@bengtson2005a],
[@udevitz2009], [@verhoef2014a], and [@southwell2008]). However, sample sizes
have often been insufficient to permit strong inference about life history
and/or seasonal variation in haul-out probabilities. For instance, Bengtson and
Cameron's [-@bengtson2004] study included 5 adult and 2 juvenile crabeater
seals, while Bengtson et al.'s [-@bengtson2005a] study was based on 6 ringed
seals in the Chukchi Sea. These studies were often further limited by logistical
constraints on fieldwork and the attachment duration or operational life
of bio-loggers. In this study, we address some of these limitations by deploying
small bio-loggers designed for longer-term attachment on a subset of the study
individuals. These devices are designed to collect data through the molt period
(when those adhered to the hair would fall off) and, in some situations, provide
multiple years of data.

@@

In this study, we used 16 years of bio-logging data to investigate the haul-out
behavior of 3 species of ice-associated seals in the Bering and Chukchi seas. We
focused our effort on bearded, ribbon, and spotted seals and did not include
ringed seals in the analysis. Ringed seals are unique in their use of snow lairs
in the spring months, and they warrant a separate study that includes additional
environmental covariates, such as the onset of snow melt and snow depth. Our
goals were threefold. First, we wished to establish baseline estimates for the
chronology of haul-out behavior in the critical spring season for each species
across different age and sex classes. Second, we refined estimates of
haul-out availability corrections for aerial surveys in order to improve
estimates of seal abundance. Previously estimated availability correction
factors (e.g., [@bengtson2005a], [@conn2014a], and [@verhoef2014a]) accounted
for variables such as the time-of-day and day-of-year, but did not investigate
meteorological variables that have been shown to influence haul-out behavior of
walruses [@reder2003; @udevitz2009]. Third, we assessed the annual variability
in haul-out timing and possible linkage to changes in the extent of seasonal sea
ice between 2005 and 2020. Our work extends the scope of previous haul-out
analyses, includes the influence of meteorological variability, and investigates
the potential impact of a changing icescape on the behavior of these species.

@@

Haul-out behavior data from `r dplyr::n_distinct(deploy_table$speno)`
bio-loggers deployed on bearded, ribbon, and spotted seals were subset to
include only records from 1 March to 15 July between 
`r min(lubridate::year(analysis_data$haulout_dt))` and 
`r max(lubridate::year(analysis_data$haulout_dt))`. Bio-loggers were of the
'SPLASH' or 'SPOT' family of tags developed by Wildlife Computers (Redmond,
Washington, USA) and either adhered to the hair on the seal or attached to the
rear flipper inter-digital webbing. The use of bio-loggers adhered to the back
or head provide some benefits over flipper mounted devices (e.g. increased
satellite transmittal rates, locations at sea) but these are lost during the
following annual molt, which, depending on deployment date, limits the duration
of haul-out data they provide. Additionally, bio-loggers attached to the head or
dorsal region are often dry while the seal is at the surface and can slightly
bias the hourly percent-dry values reported by the bio-logger. For this study,
in cases where both bio-logger types were deployed, hourly percent dry
observations from the flipper tag were preferred.

@@

```{r tblDeploy, include=TRUE, tab.id="tblDeploy",tab.cap = "Summary of bio-logger data across seal species and age classification from 1 March to 15 July 2005-2020. Total seal hours represents the sum of available data across all seals."}

@@

Haul-out behavior data was recorded and transmitted via the Argos satellite
network as hourly percent-dry timelines. For each hour of a day, the wet/dry
sensor was polled by the tag firmware every few seconds and a percent of the
hour in a dry state was calculated (Figure \@ref(fig:examplePlot)). On board the
bio-logger, hourly percent-dry data were rounded to the nearest 10 percent
inclusive of 0 and 100 percent and additional values at 3 and 98 percent. This
compression resulted in additional data transmission as each message consisted
of two complete 24-hour records. Bio-loggers were, generally, deployed and
programmed in a manner to maximize data transmission during the spring pupping
and molting period, however, hourly percent-dry data for days during a
deployment were not successfully transmitted. This is due to a variety of
factors including satellite coverage, tag availability (i.e. tags mounted to the
rear flipper are not available to the surface while the seal is at sea), tag
performance, duty cycling, and extra-terrestrial atmospheric interference.

@@

Of key interest in this study was the relationship between haul-out behavior and
weather covariates that vary with space and time. We explored the use of a
continuous-time correlated random walk [@devins.johnson2008] movement model to
predict locations at specific times. However, the sparse nature of data from
some bio-loggers, especially those mounted to the rear flipper, resulted in poor
modeling performance or convergence issues. For this study, we calculated a
weighted average daily location where the inverse of the estimated Argos or
FastLoc GPS location error was used for the weight. Each Argos location estimate
was assigned an error radius based on either the categorical location quality (
_3_=250m, _2_=500m, _1_=1500m, _0_=2500m, _A_=2500m, _B_=2500m) or, when
available, the estimated error radius from the Argos Kalman filter algorithm.
Location estimates from FastLoc GPS were all assigned an error radius of 50m.
Any days where haul-out observations were present but location data were missing
we used the last calculated weighted average daily location, and any days where
the location intersected with land were removed from the data set. We recognized
that bearded and spotted seals haul out on land. However, assessing the
relationship between haul-out behavior and weather covariates and their
availability for aerial surveys on land was outside the scope of this study.
Additionally, any daily locations on land are likely more reflective of
coordinate averaging and measurement error instead of actual use of coastal
features. Figure \@ref(fig:dataMap) shows the spatial distribution weighted
locations with available haul-out behavior data for each species across the
study area.

@@

Notably missing from this list of explanatory variables is any spatial-temporal
representation of sea ice concentration, area, or extent. This may seem
counterintuitive when modeling the haul-out behavior of seal species with such a
close association to sea ice; seals haul out in the presence of sea ice, and we
could assess the local concentration of sea ice during these events (see
[@olnes2020]). However, a major focus of this study is to develop models
applicable for aerial survey correction factors and using sea ice as a covariate
would almost certainly bias haul-out predictions towards those seals that are on
or near ice and therefore more likely to haul out. Since aerial surveys can only
detect seals on ice, abundance estimates would be missing a correction for those
seals that are away from ice (e.g., on foraging trips). Additionally, many of
the deployments consisted of a single device attached to the rear flipper of the
seal which do not provide at-sea locations, limiting our ability to fully
evaluate habitat preferences related to sea-ice. It should be noted that the
aerial survey counts are modeled with sea-ice and other habitat parameters. So,
any final abundance estimates that combine models of survey counts and our model
for proportion of seals hauled out on ice do account for the strong association
between these species and sea-ice in the Bering and Chukchi seas. Lastly, our
study was limited to the spring season when seal haul-out tendencies are
strongly influenced by pupping, nursing, breeding behavior, and molt and these
drivers are likely more influential than specific sea-ice concentration.
Crawford et al [-@crawford2019] compared haul-out probability models for ringed
seals and found those that only included season (and not sea-ice concentration)
were the most parsimonious. Further, studies to evaluate habitat selection in
relation sea-ice parameters and other environmental covariates are in progress.
For these reasons, we have elected not to use sea ice concentration as a
predictor for haul-out probability.

@@

Haul-out records for seals are often characterized by sequential hours spent
basking on ice alternating with long periods in the water (Figure
\@ref(fig:examplePlot)). Commonly used statistical models for binary data (e.g.
logistic regression) assume independence among responses, an assumption that is
clearly violated if hourly responses are modeled. Any analysis that ignores
temporal dependence in responses will thus have overstated precision
[@betts2006].

To properly account for temporal dependence within a computationally tractable
framework, we used generalized linear mixed pseudo-models (GLMPMs;
[@verhoef2010a]) to model variation in haul-out behavior as a function of (1)
covariate predictors, (2) temporally autocorrelated random effects, and (3)
individual random effects representing heterogeneity in individual behavior. We
used the {glmmLDTS} package [@verhoef2010a] to implement GLMPMs. We explored two
different model formulations for our data, and owing to the large number of
records, we fit separate models to bearded, ribbon, and spotted seal data sets.
In our first model formulation, for each species, we fitted a year-independent
model that predicted average haul-out behavior as a function of demographic,
environmental, seasonal, and diurnal effects. Second, for ribbon and spotted
seals (which had considerably more data than bearded seals), we fitted models
that included all the effects from the first model, but also permitted annual
variation in haul-out timing. This second class of models was used to examine
whether haul-out patterns varied by year and to determine the annual timing of
apparent peaks in haul-out behavior. For both models, the response variable was
the transmitted hourly percent-dry value.

@@

Unlike previous analyses of seal haul-out behavior (e.g. [@verhoef2010a],
[@conn2014a]), we also investigated the influence of sex-age class on haul-out
probabilities for all species except bearded seals because of low sample size.
While both ribbon and spotted seals exhibited a unimodal diel haul-out pattern
generally centered around local solar noon, there were key differences across
species, age, and sex that match expectations given what we know of their
ecological behavior. Ribbon seals are not known to form triads and our model
predicts a progression of increased haul-out behavior with females starting
earlier in the season than males. Notably, female ribbon seals spend a large
portion of the day in the water during the pupping period, aligning with the
hypothesis that ribbon seal females continue foraging while nursing. Spotted
seals are known to form triads during the breeding season [@burns1973] where a
female and dependent pup are accompanied on the ice by a suitor male. The male
waits for the female to wean the pup and enter estrus, and fends off any other
potential suitor males. Triad formation results in both males and females
spending a large portion of the day hauled out on ice and a protracted spring
haul-out season for both sexes. Because mating occurs in the water, females may
also be less inclined to interrupt their haul-out with foraging trips while
still nursing the pup. We see this reflected in the predicted haul-out behavior,
with both males and females exhibiting a broad distribution of time out of the
water throughout the solar day and the season. In the case of both ribbon and
spotted seals, subadults are the first to begin consistent haul-out behavior and
follow a typical phocid pattern where subadults molt first as they do not have
any reproductive constraints.

We also investigated the influence of weather on haul-out probabilities,
including wind speed, temperature, barometric pressure, precipitation, and wind
chill. These have been investigated for walruses (e.g. Udevitz et al.
[-@udevitz2009]) and a few select studies of ice-associated seals
[@hamilton2018, @perry2017]. Ribbon seals seemed to be the most influenced by
weather, with wind, temperature, barometric pressure, and precipitation all
being important components of the model. Spotted seals were most affected by
wind and barometric pressures. And, for bearded seals, the model indicated wind
and temperature had the greatest impact. In general, and as might be expected,
seals were more likely to haul-out when daily temperatures were warmer, winds
speeds were lower, barometric pressure was higher, and precipitation was lower.
These weather conditions are general indicators of increased solar radiation
which provides energetic benefits. Low winds and precipitation may also
provide an optimal situation for predator detection. Our results highlight the
importance for haul-out behavior models (and subsequent calculations of
availability during surveys) to incorporate weather covariates.

Our models detected small deviations in the timing and magnitude of annual peaks
in haul-out behavior for ribbon and spotted seals. The timing of peak haul-out
activity appears to fall within a relatively narrow time window of 3-4 weeks in
late May and early June. This consistency across 15 years is likely a reflection
of the relationship between a critical photoperiod and the timing of important
life history stages [@bronson2009; @temte1994]. However, along with a critical
photoperiod, ribbon and spotted seals are dependent upon the presence of sea ice
for pupping and molt. We didn't find any support in our models for a
relationship between the timing of peaks in haul-out behavior and the amount of
yearly maximum sea ice. This could indicate that, while the extent of spring sea
ice in the Bering sea varied widely during our study period, seals were still
able to locate sea ice and haul out. We should expect, however, that some
minimal threshold in the spatial extent or density of sea ice will have a
meaningful impact on the timing of peak haul-out behavior -- if there is no sea
ice, seals will not haul out. Additionally, while from an ecological perspective
the haul-out behavior appears consistent, the interannual differences in timing
and magnitude are large enough to have important ramifications on calculations
of abundance and trend. Those ramifications will only be exacerbated if climate
variability amplifies those interannual differences.

@@

Predictions of absolute haul-out probability in this paper were somewhat
different than those previously reported for these species, especially for
bearded seals. For instance, Ver Hoef et al. [-@verhoef2014a)] and Conn et al.
[-@conn2014a] used haul-out correction factors with maximums of 0.66 for bearded
seals, 0.62 for ribbon seals, and 0.54 for spotted seals, where maximums
corresponded to times near solar noon in mid-late May. Applying models that
ignore age, sex, and year effects, these probabilities were 0.38, 0.72, and
0.60, respectively, under the current analysis framework. Our current estimates
estimates of haul-out probability reflect increased sample sizes in terms of
number of animals, but also improvements to the way data were prepared prior to
analysis.

@@

We recognize that the species and ecosystems studied in this paper are within
the traditional range of the Inpuiaq and Yupik people. And, we would like to
acknowledge and thank them as the original stewards of these ecosystems. The
deployment of bio-logging devices used in this study were often done in
collaboration with Alaska Native seal hungers and their communities. We would
like to especially acknowledge the communities of Kotzebue, Koyuk, Utqiaġvik,
and Ulguniq (Wainwright) and the following individuals: James Adams, Jeff
Barger, David Barr, Wendell Booth, Cyrus Harris, Nereus 'Doc' Harris, Grover
Harris, Lee Harris, Tom Jones, Frank Garfield, Brenda Goodwin, Henry Goodwin,
John Goodwin, Pearl Goodwin, Willie Goodwin, Brett Kirk, Noah Naylor, Virgil
Naylor Jr., Virgil Naylor Sr., Dan Savetilik, Allen Stone, and Randy Toshavik
from Kotzebue, Alaska; Merlin Henry from Koyuk, Alaska; Billy Adams, Mary-Ellen
Ahmaogak, James Aiken, Tim Aiken, Sarah Coburn, Jason Herreman, Howard Kittick,
Gilbert Leavitt, Isaac Leavitt, J.R. Leavitt, Enoch Oktollik, Shawn Oktollik,
Stacey Osborn, Fred Rexford, Chuck Schaeffer, Ross Schaeffer, Bob Shears, and
Joe Skin from the North Slope Borough, Alaska;

@@

Telemetry data from the Alaska Department of Fish and Game (ADFG) and the North
Slope Borough Department of Wildlife Management (NSB) were important
contributions to the findings presented here. Both agencies collaborate
closely with Alaska Native communities and the contributions of local
hunters to the successful capture, deployment, and understanding of seal
ecology are invaluable. Deployments in the western Bering Sea and Sea of
Okhotsk were done in collaboration with Russian colleagues and North
Pacific Wildlife.

>>>>

For ice-associated seals, haul-out behavior is tightly linked with sea ice.
Ribbon seals (_Histriophoca fasciata_) haul out of the water almost exclusively
on sea ice and are mostly pelagic outside the spring pupping, breeding, and
molting season [@boveng2018a]. While spotted (_Phoca largha_) and bearded
(_Erignathus barbatus_) seals rest on coastal features, they strongly prefer sea
ice as a resting platform during the spring and early summer [@frost2018].
Ringed seals (_Phoca hispida_) --- not included in this study --- bask on
sea-ice but also within snow lairs built during the spring. The remoteness of
the Bering and Chukchi seas mean direct scientific observation of seal behavior
is impractical. Thus, bio-logging devices are especially useful tools for
collecting key information on movement and haul-out behavior for these species.

@@

Knowledge of haul-out patterns is not only important for understanding natural
history and ecology, but also for developing "availability" correction factors
for aerial surveys. Specifically, researchers need to know the fraction of seals
hauled out (versus in the water) when aerial surveys are conducted. Studies
estimating availability correction factors for seals typically use logistic
regression-style analyses to estimate the time-specific probability of being
hauled out based on 'wet/dry' data relayed by bio-loggers. In these models,
haul-out probabilities were expressed as a function of predictive covariates,
such as time-of-day, day-of-year, sex, age class, and environmental conditions
(e.g., [@reder2003], [@bengtson2004], [@bengtson2005a], [@udevitz2009],
[@verhoef2014a], and [@southwell2008]). However, sample sizes have often been
insufficient to permit strong inference about life history and/or seasonal
variation in haul-out probabilities. For instance, Bengtson and Cameron's
[-@bengtson2004] study included 5 adult and 2 juvenile crabeater seals, while
Bengtson et al.'s [-@bengtson2005a] study was based on 6 ringed seals in the
Chukchi Sea. These studies were often further limited by logistical constraints
on fieldwork and the attachment duration or operational life of bio-loggers. In
this study, we address some of these limitations by deploying small bio-loggers
designed for longer-term attachment on rear flippers of a subset of the study
individuals. These devices are designed to collect data through the molt period
(when those adhered to the hair would fall off) and, in some situations, provide
multiple years of data.

@@

In this study, we used 16 years of bio-logging data to investigate the haul-out
behavior of 3 species of ice-associated seals in the Bering and Chukchi seas. We
focused our effort on bearded, ribbon, and spotted seals. Our
goals were threefold. First, we wished to establish baseline estimates for the
chronology of haul-out behavior in the critical spring season for each species
across different age and sex classes. Second, we refined estimates of
haul-out availability corrections for aerial surveys in order to improve
estimates of seal abundance. Previously estimated availability correction
factors (e.g., [@bengtson2005a], [@conn2014a], and [@verhoef2014a]) accounted
for variables such as the time-of-day and day-of-year, but did not investigate
meteorological variables that have been shown to influence haul-out behavior of
walruses [@reder2003; @udevitz2009]. Third, we assessed the annual variability
in haul-out timing and possible linkage to changes in the extent of seasonal sea
ice between 2005 and 2020. Our work extends the scope of previous haul-out
analyses, includes the influence of meteorological variability, and investigates
the potential impact of a changing icescape on the behavior of these species.

@@

Haul-out behavior data from bio-loggers deployed on 
`r dplyr::n_distinct(deploy_table$speno)` bearded, ribbon, and spotted seals were
subset to include only records from 1 March to 15 July between 
`r min(lubridate::year(analysis_data$haulout_dt))` and 
`r max(lubridate::year(analysis_data$haulout_dt))`. Bio-loggers were of the
'SPLASH' or 'SPOT' family of tags developed by Wildlife Computers (Redmond,
Washington, USA) and either adhered to the hair on the seal or attached to the
rear flipper inter-digital webbing. The use of bio-loggers adhered to the back
or head provide some benefits over flipper mounted devices (e.g. increased
satellite transmittal rates, locations at sea) but these are lost during the
following annual molt, which, depending on deployment date, limits the duration
of haul-out data they provide. Additionally, bio-loggers attached to the head or
dorsal region are often dry while the seal is at the surface and can slightly
bias the hourly percent-dry values reported by the bio-logger. For this study,
in cases where both bio-logger types were deployed, hourly percent dry
observations from the flipper tag were preferred.

@@

```{r tblDeploy, include=TRUE, tab.id="tblDeploy",tab.cap = "Summary of bio-logger data across seal species and age classification from 1 March to 15 July 2005-2020. Total seal hours represents the sum of available data across all seals. Because young-of-the-year are advanced to subadult on 1 March of the following year, some individual seals are represented in both columns in this table"}

@@

Haul-out behavior data were recorded and transmitted via the Argos satellite
network as hourly percent-dry timelines. For each hour of a day, the wet/dry
sensor was polled by the tag firmware every few seconds and a percent of the
hour in a dry state was calculated (Figure \@ref(fig:examplePlot)). On board the
bio-logger, hourly percent-dry data were rounded to the nearest 10 percent
inclusive of 0 and 100 percent and additional values at 3 and 98 percent. This
compression resulted in additional data transmission as each message consisted
of two complete 24-hour records. Bio-loggers were, generally, deployed and
programmed in a manner to maximize data transmission during the spring pupping
and molting period; however, hourly percent-dry data were not always
successfully transmitted. This is due to a variety of factors including
satellite coverage, tag availability (i.e. tags mounted to the rear flipper
often do not transmit while at sea), tag performance, duty cycling, and
extra-terrestrial atmospheric interference. Fortunately, missing records do not
seem to greatly bias inference about haul-out probabilities [@conn2012b].

@@

Of key interest in this study was the relationship between haul-out behavior and
weather covariates that vary with time and animal location. We explored the use
of a continuous-time correlated random walk [@devins.johnson2008] movement model
to predict locations at specific times. However, the sparse nature of data from
some bio-loggers, especially those mounted to the rear flipper, resulted in poor
modeling performance or convergence issues. For this study, we calculated a
weighted average daily location where the inverse of the estimated Argos or
FastLoc GPS location error was used for the weight. Each Argos location estimate
was assigned an error radius based on either the categorical location quality (
_3_=250m, _2_=500m, _1_=1500m, _0_=2500m [@lopez2013]; we chose 2500m for
location classes _A_ and _B_) or, when available, the estimated error radius
from the Argos Kalman filter algorithm. Location estimates from FastLoc GPS were
all assigned an error radius of 50m. Any days where haul-out observations were
present but location data were missing we used the last calculated weighted
average daily location, and any days where the location intersected with land
were removed from the data set. We recognized that bearded and spotted seals
haul out on land. However, assessing the relationship between haul-out behavior
and weather covariates and their availability for aerial surveys on land was
outside the scope of this study. Additionally, any daily locations on land are
likely more reflective of coordinate averaging and measurement error instead of
actual use of coastal features. Figure \@ref(fig:dataMap) shows the spatial
distribution weighted locations with available haul-out behavior data for each
species across the study area.

@@

Notably missing from this list of explanatory variables is any spatial-temporal
representation of sea ice concentration, area, or extent. This may seem
counterintuitive when modeling the haul-out behavior of seal species with such a
close association to sea ice; seals haul out in the presence of sea ice, and we
could assess the local concentration of sea ice during these events (see
[@olnes2020]). However, a major focus of this study is to develop models
applicable for aerial survey correction factors and using sea ice as a covariate
would almost certainly bias haul-out predictions towards those seals that are on
or near ice and therefore more likely to haul out. Since aerial surveys can only
detect seals on ice, abundance estimates would be missing a correction for those
seals that are away from ice (e.g., on foraging trips). Additionally, many of
the deployments consisted of a single device attached to the rear flipper of the
seal which do not provide at-sea locations, limiting our ability to fully
evaluate fine-scale habitat preferences related to sea-ice. Lastly, our study
was limited to the spring season when seal haul-out tendencies are strongly
influenced by pupping, nursing, breeding behavior, and molt and these drivers
are likely more influential than specific sea-ice concentration. Crawford et al
[-@crawford2019] compared haul-out probability models for ringed seals and found
those that only included season (and not sea-ice concentration) were the most
parsimonious. For these reasons, we have elected not to use sea ice
concentration as a predictor for haul-out probability in the present study
(note, however, that habitat selection analyses incorporating sea-ice
concentration are a focus of current and future research).

@@

Haul-out records for seals are often characterized by sequential hours spent
basking on ice alternating with long periods in the water (Figure
\@ref(fig:examplePlot)). Commonly used statistical models for binary data (e.g.
logistic regression) assume independence among responses, an assumption that is
clearly violated if hourly responses are modeled. Any analysis that ignores
temporal autocorrelation in responses will thus have overstated precision
[@betts2006].

To properly account for temporal dependence within a computationally tractable
framework, we used generalized linear mixed pseudo-models (GLMPMs;
[@verhoef2010a]) to model variation in haul-out behavior as a function of (1)
covariate predictors, (2) temporally autocorrelated random effects, and (3)
individual random effects representing heterogeneity in individual behavior. We
used the glmmLDTS package [@verhoef2010a] to implement GLMPMs. We explored two
different model formulations for our data, and owing to the large number of
records, we fit separate models to bearded, ribbon, and spotted seal data sets.
In our first model formulation, for each species, we fitted a year-independent
model that predicted average haul-out behavior as a function of demographic,
environmental, seasonal, and diurnal effects. Second, for ribbon and spotted
seals (which had considerably more data than bearded seals), we fitted models
that included all the effects from the first model, but also permitted annual
variation in haul-out timing. This second class of models was used to examine
whether haul-out patterns varied by year and to determine the annual timing of
apparent peaks in haul-out behavior. For both models, the response variable was
the transmitted hourly percent-dry value.

@@

Unlike previous analyses of seal haul-out behavior (e.g. [@verhoef2010a],
[@conn2014a]), we also investigated the influence of sex-age class on haul-out
probabilities for all species except bearded seals because of low sample size.
Although ribbon and spotted seals exhibited a unimodal diel haul-out pattern
generally centered around local solar noon, there were key differences across
species, age, and sex that match expectations given what we know of their
ecological behavior. Spotted seals are known to form triads during the breeding
season [@burns1973] where a female and dependent pup are accompanied on the ice
by a suitor male. The male waits for the female to wean the pup and enter
estrus, and fends off any other potential suitor males. Triad formation results
in both males and females spending a large portion of the day hauled out on ice
and a protracted spring haul-out season for both sexes. Because mating occurs in
the water, females may also be less inclined to interrupt their haul-out with
foraging trips while still nursing the pup. We see this reflected in the
predicted haul-out behavior, with both males and females exhibiting a broad
distribution of time out of the water throughout the solar day and the
season.Ribbon seals are not known to form triads and our model predicts a
progression of increased haul-out behavior with females starting earlier in the
season than males. Notably, female ribbon seals spend a large portion of the day
in the water during the pupping period, aligning with the hypothesis that ribbon
seal females continue foraging while nursing.  In the case of both ribbon and
spotted seals, subadults are the first to begin consistent haul-out behavior and
follow a typical phocid pattern where subadults molt first as they do not have
any reproductive constraints.

We also investigated the influence of weather on haul-out probabilities,
including wind speed, temperature, barometric pressure, precipitation, and wind
chill. These have been investigated for walruses (e.g. Udevitz et al.
[-@udevitz2009]) and a few select studies of ice-associated seals
[@hamilton2018, @perry2017]. Ribbon seals seemed to be the most influenced by
weather, with wind, temperature, barometric pressure, and precipitation all
being important components of the model. Spotted seals were most affected by
wind and barometric pressure. And, for bearded seals, the model indicated wind
and temperature had the greatest impact. In general, and as might be expected,
seals were more likely to haul-out when daily temperatures were warmer, winds
speeds were lower, barometric pressure was higher, and precipitation was lower.
These weather conditions are general indicators of increased solar radiation
which provides energetic benefits. Low winds and precipitation may also
provide an optimal situation for predator detection. Our results highlight the
importance of incorporating weather covariates when analyzing haul-out 
behavior and calculating availability corrections for aerial surveys.

Our models detected small deviations in the timing and magnitude of annual peaks
in haul-out behavior for ribbon and spotted seals. The timing of peak haul-out
activity appears to fall within a relatively narrow time window of 3-4 weeks in
late May and early June. This consistency across 15 years is likely a reflection
of the relationship between a critical photoperiod and the timing of important
life history stages [@bronson2009; @temte1994]. However, along with a critical
photoperiod, ribbon and spotted seals are dependent upon the presence of sea ice
for pupping and molt. We didn't find any support in our models for a
relationship between the timing of peaks in haul-out behavior and the amount of
yearly maximum sea ice. This could indicate that, while the extent of spring sea
ice in the Bering sea varied widely during our study period, seals were still
able to locate sea ice and haul out. We should expect, however, that some
minimal threshold in the spatial extent or density of sea ice will have a
meaningful impact on the timing of peak haul-out behavior -- if there is no sea
ice, seals will not haul out. Additionally, while from an ecological perspective
the haul-out behavior appears consistent, the interannual differences in timing
and magnitude are large enough to have important ramifications on calculations
of abundance and trend. Those ramifications will only be exacerbated if climate
variability amplifies interannual differences.

@@

Predictions of absolute haul-out probability in this paper were somewhat
different than those previously reported for these species, especially for
bearded seals. For instance, Ver Hoef et al. [-@verhoef2014a)] and Conn et al.
[-@conn2014a] used haul-out correction factors with maximums of 0.66 for bearded
seals, 0.62 for ribbon seals, and 0.54 for spotted seals, where maximums
corresponded to times near solar noon in mid-late May. Applying models that
ignore age, sex, and year effects, these probabilities were 0.38, 0.72, and
0.60, respectively, under the current analysis framework. Our current estimates
of haul-out probability reflect increased sample sizes in terms of
number of animals, but also improvements to the way data were prepared prior to
analysis.

@@

We recognize that the species and ecosystems studied in this paper are within
the traditional range of the Inpuiaq and Yupik people. And, we would like to
acknowledge and thank them as the original stewards of these ecosystems. The
deployment of bio-logging devices used in this study were often done in
collaboration with Alaska Native seal hunters and their communities. We would
like to especially acknowledge the communities of Kotzebue, Koyuk, Utqiaġvik,
and Ulguniq (Wainwright) and the following individuals: James Adams, Jeff
Barger, David Barr, Wendell Booth, Cyrus Harris, Nereus 'Doc' Harris, Grover
Harris, Lee Harris, Tom Jones, Frank Garfield, Brenda Goodwin, Henry Goodwin,
John Goodwin, Pearl Goodwin, Willie Goodwin, Brett Kirk, Noah Naylor, Virgil
Naylor Jr., Virgil Naylor Sr., Dan Savetilik, Allen Stone, and Randy Toshavik
from Kotzebue, Alaska; Merlin Henry from Koyuk, Alaska; Billy Adams, Mary-Ellen
Ahmaogak, James Aiken, Tim Aiken, Sarah Coburn, Jason Herreman, Howard Kittick,
Gilbert Leavitt, Isaac Leavitt, J.R. Leavitt, Enoch Oktollik, Shawn Oktollik,
Stacey Osborn, Fred Rexford, Chuck Schaeffer, Ross Schaeffer, Bob Shears, and
Joe Skin from the North Slope Borough, Alaska;

@@

Telemetry data from the Alaska Department of Fish and Game (ADFG) and the North
Slope Borough Department of Wildlife Management (NSB) were important
contributions to the findings presented here. Both agencies collaborate closely
with Alaska Native communities and the contributions of local hunters to the
successful capture, deployment, and understanding of seal ecology are
invaluable. Deployments in the western Bering Sea were done in collaboration
with Russian colleagues and North Pacific Wildlife.
